---
title: "Mass 211: Model Selection"
output: html_notebook
---

## Distribution of Counts

```{r}
dat <- count.town

ggplot(dat, aes(x = p_call, y = ..count..)) +
  geom_histogram()
```



## Poverty

Linear regression (OLS, ordinary least square):

```{r}
dat <- zip.geo@data
summary(m1 <- lm(
  p_call ~
    White + Black + Hispanic + Asian + EthHet +
    MedHouseIncome + GINI + PubAssist + UnempRate + BelowPoverty +
    PopDen + p.urbanpop +
    AgeU5 + Age1824 + Age2534 + Age3544 + Age4554 +
    AtLeastBachelor,
  data = dat))
```

### Generaized Linear Model - Negative Binomial

This model predicts the log value of the expected number of calls. Log value of the population is used as an offset.

#### Town Level

After eliminating extreme values, and using Backward Reduction, the model is like:

```{r}
dat <- town.geo@data %>%
  # remove extrem values
  filter(p_call < quantile(p_call, .99, na.rm=TRUE)) %>%
  mutate(
    shape_area = as.integer(shape_area),
    TotalPop = as.integer(TotalPop)
  ) %>%
  # count numbers are integers
  mutate_at(vars(starts_with('n_')), as.integer) %>%
  # Standardize all variables
  mutate_if(is.double, function(x) scale(x)[,1])

summary(m1 <- MASS::glm.nb(
  n_call ~
    TwoOrMore + EthHet +
    MedHouseIncome + 
    Age1824 + 
    I(Age6061 + Age6264 + Age6574) +
    AtLeastBachelor + offset(log(TotalPop)),
  data = dat
))
```

#### ZIP Code Level

```{r}
dat <- zip.geo@data %>%
  # remove extrem values
  filter(p_call < quantile(p_call, .99, na.rm=TRUE)) %>%
  mutate(
    shape_area = as.integer(shape_area),
    TotalPop = as.integer(TotalPop)
  ) %>%
  # count numbers are integers
  mutate_at(vars(starts_with('n_')), as.integer) %>%
  # Standardize all variables
  mutate_if(is.double, function(x) scale(x)[,1])

summary(m1 <- MASS::glm.nb(
  n_call ~
    White + Black + Hispanic + Asian + TwoOrMore + EthHet +
    MedHouseIncome + GINI + PubAssist + UnempRate + BelowPoverty +
    PopDen + p.urban_pop +
    AgeU5 + Age1824 + Age2534 + Age3544 + Age4554 +
    Age5559 + I(Age6061 + Age6264 + Age6574) +
    AtLeastBachelor +
    offset(log(TotalPop)),
  data = dat
))
```

```{r}

```

```{r}
library(mvabund)
summary(m1 <- manyglm(
  n_call ~
    Black +  Asian + EthHet +
    MedHouseIncome + GINI + PubAssist +
    PopDen  +
    AtLeastBachelor +
    # offset
    offset(log(TotalPop)),
  data = dat,
  family = 'poisson'
))
plot(m1)
```
```{r}
summary(m1 <- glm(
  n_call ~
    Black +  Asian + EthHet +
    MedHouseIncome + GINI + PubAssist +
    PopDen  +
    AtLeastBachelor +
    # offset
    offset(log(TotalPop)),
  data = dat,
  family = 'poisson'
))
```

```{r}
summary(m1 <- glm(
  n_call ~
    Black + 
    EthHet +
    MedHouseIncome +
    AtLeastBachelor +
    # offset
    offset(log(TotalPop)),
  data = dat,
  family = 'poisson'
))
```

```{r}
anova(m1, test="Chisq")
```

```{r}
plot(m1, which=1)
```

```{r}
summary(m2 <- MASS::glm.nb(
  n_call ~
    Black + 
    EthHet +
    MedHouseIncome +
    AtLeastBachelor +
    # offset
    offset(log(TotalPop)),
  data = dat
))
```

```{r}
plot(m2, which=1)
```

```{r}
anova(m1, test="Chisq")
```
```
```{r}
plot(m1, which=1)
```


When reducing varibles using 

```{r}
summary(m1 <- MASS::glm.nb(
  n_call ~
    White + Hispanic + Asian + TwoOrMore + EthHet +
    MedHouseIncome + PubAssist + 
    p.urban_pop +
    Age1824 + 
    I(Age6061 + Age6264 + Age6574) +
    offset(log(TotalPop)),
  data = dat
))
```


County

```{r}
```


### Zero-Inflated Poisson model

```{r}
dat <- zip.geo@data %>%
  mutate(
    MedHouseIncome = scale(MedHouseIncome)[,1],
    PopDen = scale(PopDen)[,1]
  )

AddResidual <- function(dat) {
  zeroinfl <- pscl::zeroinfl
  summary(m1 <- zeroinfl(
    n_call_childcare ~
      Black + EthHet +
      MedHouseIncome + PubAssist + UnempRate +
      PopDen + p.urbanpop +
      Age3544 +
      I(Age6061 + Age6264 + Age6574) +
      offset(log(TotalPop))
       | offset(log(TotalPop)) + MedHouseIncome + p.urbanpop,
    dist = "negbin",
    link = "logit",
    data = dat
  ))
  dat <- dat %>% mutate(residual = NA) %>% select(name, residual)
  dat[names(m1$residuals), 'residual'] <- m1$residuals
  dat$residual
}

zip.geo@data$residual <- dat$residual
```

### Residules on a map

```{r}

```



## Public Health

```{r}
zeroinfl <- pscl::zeroinfl
summary(m1 <- zeroinfl(
  n_call ~
    Black + EthHet +
    MedHouseIncome + PubAssist + UnempRate +
    PopDen + p.urbanpop +
    Age3544 +
    I(Age6061 + Age6264 + Age6574)
     | TotalPop,
  dist = "negbin",
  link = "logit",
  data = dat
))
```
